A simplified dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-01-11 JTB_ms/JTB_ms.tex

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states,
corresponding to either positive or negative growth (Fig. 2). The
mechanism that produces these alternate stable states is the positive
feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), the system cannot
escape this positive feedback and cannot grow. However, if symbiont
biomass is initially high (i.e., a ``healthy'' coral), then the system
remains in a nitrogen-limited state with positive growth. For practical
purposes, this section of the manuscript considers only positive growth
steady states under constant environments; the subsequent ``dynamic
behavior'' section explores how changing environments may cause the
system to switch between alternate stable states, which we interpret in
the context of coral bleaching.

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig3.png}
\caption{Steady state values of \textbf{A)} specific growth (Cmol
Cmol\textsuperscript{-1} d\textsuperscript{-1}), \textbf{B)} the
symbiont to host biomass ratio (CmolS CmolH\textsuperscript{-1}),
\textbf{C)} relative C- or N-limitation of host biomass formation, and
\textbf{D)} relative CO\textsubscript{2}- or light-limitation of
symbiont photosynthesis, across gradients of external irradiance and
dissolved inorganic nitrogen. Note that typical conditions for reefs are
\textasciitilde{}1e-7 M DIN and 10-20 mol photons m\textsuperscript{-2}
d\textsuperscript{-1}. Simulations for each combination of light and
nutrients (41 points along each axis) were run for 100 days with a time
step of 1 day. Negative steady state growth rates, and corresponding S:H
ratios, were set to zero.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig5.png}
\caption{Light-driven seasonal dynamics of symbiont abundance and coral
growth. Light input (upper panel) was designed as a sinusoidal curve
with a period of one year, with maximum and minimum values of 44 and 22
mol photons m\textsuperscript{-2} d\textsuperscript{-1}, based on
Stimson (1997). The dynamic behavior of symiont to host biomass ratio
(middle panel) and the specific growth rate of host biomass (lower
panel) show seasonal oscillations that are greater in magnitude under
higher nutrients (1e-5 molN L\textsuperscript{-1}; dashed lines)
relative to lower nutrients (1e-7 molN L\textsuperscript{-1}; solid
lines), consistent with the findings of Stimson (1997). Prey density was
set at 1e-5 CmolX L\textsuperscript{-1}.}
\end{figure}

@@

The importance of host CCM activity establishes significant interactive
roles for other factors in influencing coral bleaching responses. For
example, simulations of high light stress (Fig. 7) demonstrate that
bleaching can be attenuated by heterotrophic feeding, a phenomenon which
has been observed experimentally (Borell et al. 2008). The mechanism
underlying this prediction is that feeding by the host increases host
CCM activity, which delays the onset of CO\textsubscript{2}-limitation
of photosynthesis and reduces bleaching severity. On the other hand,
elevated nutrients exacerbate bleaching (Fig. 7), since higher symbiont
densities are more susceptible to CO\textsubscript{2}-limitation
(Wooldridge 2009). Several experimental (Cunning and Baker 2013; Vega
Thurber et al. 2014) and correlational studies (Wooldridge and Done
2009) are consistent with this mechanistic link between high nutrients
and bleaching.

@@

Since bleaching is induced by an external stressor, recovery from
bleaching cannot occur unless the stressor is alleviated. In natural
settings, this typically occurs by seasonal declines in temperature and
light. However, alternate stable states lead to significant hysteresis,
indicating that a coral cannot recover along the same trajectory it
followed during bleaching; indeed, the stressor must be alleviated well
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 2; Fig. X). This is because under the same
external conditions, a bleached coral (relative to a healthy one) is
characterized by greater light amplification and weaker CCM activity,
which serve to maintain the bleached state. The magnitude of hysteresis
will depend on external nutrients and prey availability, as well as
model parameters, yet the core behavior of hysteresis illustrates the
alternate stable states that exist in the symbiotic system.

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Recovery from bleaching under varying N:C availability. Higher
N:C ratios in a heterotrophic food source (effectively representing
higher external DIN and/or lower feeding rates) cause a larger overshoot
in the S:H biomass ratio, and prolong the duration of time until the
system `recovers' by re-establishing nitrogen-limitation (represented by
black dots on each trajectory). All simulations run with default
parameters (except varying nNX), L=10 mol photons m\textsuperscript{2}
d\textsuperscript{-1}, DIN=1e-7 molN L\textsuperscript{-1}, prey=1e-6
CmolX L\textsuperscript{-1}, and initial S:H biomass=0.0001.}
\end{figure}

The dynamics of recovery from bleaching reveal another interesting
behavior of the system: under some conditions, the S:H biomass ratio may
increase to a higher value that `overshoots' the ratio maintained in the
`healthy' state under the same conditions (Fig. 8; Fig. S9). In fact,
unusually high symbiont densities have been observed after recovery from
bleaching in both experimental (Cunning, Silverstein, and Baker 2015)
and field studies (Kemp et al. 2014), and have been interpreted as a
potential `disequilibrium in host-symbiont regulation' (Kemp et al.
2014). The model explains this `overshoot' as an approach to a
bifurcation point in the dynamical system \textless{}\textgreater{}. As
the symbiont begins repopulating the host, CO\textsubscript{2} is
limiting and nitrogen is replete, allowing nearly all fixed carbon to be
used for symbiont growth. As a result, host growth remains strongly
carbon-limited, CCM activity remains weak, and photosynthesis remains
CO\textsubscript{2}-limited. As long as this carbon-limitation persists,
S:H biomass will continue to increase beyond the `healthy' steady state
level, until the symbiont population eventually reaches a high enough
density such that nitrogen limits its growth more strongly than carbon.
At this point, which represents the peak of the `overshoot', the
symbiont begins to share carbon with the host, which then supplies more
CO\textsubscript{2} for photosynthesis, etc., causing the system to
rapidly escape the positive feedbacks that were maintaining
carbon-limitation (see Fig. S9 for illustration of all fluxes during
overshoot). The re-establishment of nitrogen-limitation resumes high
rates of carbon fixation and coral growth, bringing the S:H biomass
ratio back down to a stable equilibrium.

The occurrence and the magnitude of this overshoot depend on the
relative inputs of carbon and nitrogen to the system -- any factor that
enhances carbon-limitation (e.g.~high DIN and/or low feeding) therefore
magnifies the overshoot and prolongs the dysfunctional, carbon-limited
state of the symbiosis (Fig. 8). On the other hand, factors that favor
nitrogen-limitation, such as low external DIN and/or high feeding rates,
will accelerate the re-establishment of nitrogen-limitation and prevent
an overshoot from occurring at all. While many scenarios are possible
under different environmental conditions, we illustrate the general
effects of varying N and C availability on recovery from bleaching with
a series of simulations that vary the N:C ratio of host's heterotrophic
food source (Fig. 8): lower N:C ratios (effectively representing lower
DIN and/or higher heterotrophy) favor nitrogen-limitation and more rapid
recovery, while higher N:C ratios (effectively representing higher DIN
and/or lower heterotrophy) favor carbon-limitation and prolonged
recovery with a larger overshoot in S:H biomass.

Importantly, the re-establishment of nitrogen-limitation is the most
important diagnostic of recovery to a `healthy' state, as this is when
high carbon fixation and growth rates are resumed. A high S:H biomass
ratio alone does not indicate that a `healthy' state has been reached,
since overall C-limitation may still persist (e.g., high N:C ratios in
Fig. 8). This could explain why corals that have apparently recovered
their symbiont populations after bleaching still have reduced growth and
reproduction rates until much later (xxx, Hughes). Therefore,
acquisition of carbon from a source other than the symbiont may be
extremely important for the host in order to recover from bleaching.
Indeed, feeding has been shown to promote more rapid recovery from
bleaching (xxx). Moreover, if the host obtains carbon from another
independent source, such as direct uptake of DOC (xxx), this could be
even more important in ensuring rapid recovery from a bleached state.

@@

@@

@@

@@

>>>>

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states,
corresponding to either positive or negative growth (Fig. 2). The
mechanism that produces these alternate stable states is the positive
feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), the system cannot
escape this positive feedback and cannot grow. However, if symbiont
biomass is initially high (i.e., a ``healthy'' coral), then the system
remains in a nitrogen-limited state with positive growth. For practical
purposes, this section of the manuscript considers only positive growth
steady states under constant environments; subsequently, we explore how
environmental forcing may cause the system to switch between alternate
stable states, which we interpret in the context of coral bleaching (see
``Coral Bleaching and Recovery'', below).

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig3.png}
\caption{Steady state values of (A) specific growth (Cmol
Cmol\textsuperscript{-1} d\textsuperscript{-1}) and (B) the symbiont to
host biomass ratio (CmolS CmolH\textsuperscript{-1}) across gradients of
external irradiance and dissolved inorganic nitrogen. Note that typical
conditions for reefs are \textasciitilde{}1e-7 M DIN and 10-20 mol
photons m\textsuperscript{-2} d\textsuperscript{-1}. Simulations for
each combination of light and nutrients (41 points along each axis) were
run for 100 days with a time step of 1 day. All parameters were set to
default values and prey density set to zero. Negative steady state
growth rates, and corresponding S:H ratios, were set to zero.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig5.png}
\caption{Light-driven seasonal dynamics of symbiont abundance and coral
growth. Light input (upper panel) was designed as a sinusoidal curve
with a period of one year, with maximum and minimum values of 44 and 22
mol photons m\textsuperscript{-2} d\textsuperscript{-1}, based on
Stimson (1997). The dynamic behavior of symbiont to host biomass ratio
(middle panel) and the specific growth rate of host biomass (lower
panel) show seasonal oscillations that are greater in magnitude under
higher nutrients (1e-5 molN L\textsuperscript{-1}; dashed lines)
relative to lower nutrients (1e-7 molN L\textsuperscript{-1}; solid
lines), consistent with the findings of Stimson (1997). Prey density was
set at 1e-5 CmolX L\textsuperscript{-1}.}
\end{figure}

@@

The importance of host CCM activity establishes significant interactive
roles for other factors in influencing coral bleaching responses. For
example, simulations of high light stress (Fig. 7) demonstrate that
bleaching can be attenuated by heterotrophic feeding, a phenomenon which
has been observed experimentally (Borell et al. 2008). The mechanism
underlying this prediction is that feeding by the host increases host
CCM activity, which delays the onset of CO\textsubscript{2}-limitation
of photosynthesis and reduces bleaching severity. On the other hand,
elevated nutrients exacerbate bleaching (Fig. 7), since higher symbiont
densities are more susceptible to CO\textsubscript{2}-limitation
(Wooldridge 2009). Several experimental (Wiedenmann et al. 2013; Cunning
and Baker 2013; Vega Thurber et al. 2014) and correlational studies
(Wooldridge and Done 2009) are consistent with this mechanistic link
between high nutrients and bleaching.

@@

Since bleaching can be understood as a transition from a
nitrogen-limited state with high S:H biomass to a carbon-limited state
with low S:H biomass, induced by an external stressor, recovery can be
understood as a switch back to the nitrogen-limited state once the
external stressor is alleviated. In natural settings, this typically
occurs through seasonal declines in temperature and light. However,
hysteresis associated with the system's alternate stable states
indicates that the symbiosis cannot recover along the same trajectory it
followed during bleaching; indeed, the stressor must be alleviated
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 2). This is because under the same external
conditions, a bleached coral with low S:H biomass (relative to a healthy
coral with high S:H biomass) is characterized by greater light
amplification and weaker CCM activity, which enhance photooxidative
stress and serve to maintain the carbon-limited state. In order for the
system to recover, the stressor must be reduced enough such that
photooxidative stress ceases and translocation of carbon from symbiont
to host is resumed. Once this occurs, the host can energize its CCMs,
which further enhances carbon fixation and accelerates the system back
toward a nitrogen-limited state, indicative of recovery. The conditions
under which recovery can occur -- which determine the magnitude of
hysteresis (Fig. SX) -- depend on the relative abundance of nitrogen and
carbon in the environment. Higher food levels, representing a
non-autotrophic carbon source for the host, make it easier for the host
to overcome carbon-limitation, thus providing a potential mechanism
underlying observations that feeding aids recovery from bleaching
{[}xxx{]}. Conversely, high external DIN impedes the re-establishment of
nitrogen-limitation, making recovery from bleaching more difficult
(i.e., narrowing the sets of conditions under which recovery is
possible).

Dynamic simulations of recovery reveal another interesting behavior of
the system: under some conditions, an `overshoot' occurs in which S:H
biomass temporarily increases beyond the ratio maintained in the
`healthy' state, before returning to stabilize at this value (Fig. 8;
Fig. S9). In fact, unusually high symbiont densities have been observed
following bleaching in both experimental (Cunning, Silverstein, and
Baker 2015) and field studies (Kemp et al. 2014), and have been
interpreted as a potential `disequilibrium in host-symbiont regulation'
(Kemp et al. 2014). The model reveals the dynamics of this `overshoot'
as follows: as symbionts repopulate the host, photosynthesis becomes
increasingly CO\textsubscript{2}-limited due to weak CCM activity of the
carbon-limited host. Thus, although symbiont growth is not yet
carbon-limited, a growing symbiont population has less and less excess
carbon (per symbiont) to share, and is thus moving toward
carbon-limitation. Meanwhile, because S:H biomass is increasing, the
host receives more and more carbon per unit host biomass, and is thus
moving away from carbon-limitation. If the host overcomes
carbon-limitation \emph{before} the symbiont reaches it, then the system
rapidly transitions to the nitrogen-limited state and the S:H ratio
stabilizes without an overshoot (Fig. 8A). However, if the symbiont
becomes carbon-limited first (Fig. 8B, 8C), then carbon translocation
per symbiont declines further and photosynthesis becomes more
CO\textsubscript{2}-limited, which maintains carbon-limitation of the
host. In this situation, continued growth of less and less productive
symbionts causes S:H biomass to reach a much higher level before the
host finally overcomes carbon-limitation. At this point, representing
the peak of the overshoot, the transition to nitrogen-limitation finally
takes place and the S:H biomass ratio declines and stabilizes.

Whether this `overshoot' occurs or not represents a bifurcation point in
the dynamical system that is defined by the relative availability of
carbon and nitrogen to the host -- any factor that enhances
carbon-limitation of host growth (e.g.~high DIN and/or low feeding)
therefore magnifies the overshoot and prolongs the dysfunctional,
carbon-limited state of the symbiosis (Fig. 8). On the other hand,
factors that favor nitrogen-limitation, such as low external DIN and/or
high feeding rates, will accelerate the re-establishment of
nitrogen-limitation and prevent an overshoot from occurring at all.
While many scenarios are possible under different environmental
conditions, we illustrate the general effects of varying N and C
availability on recovery from bleaching with a series of simulations
that vary the N:C ratio of host's heterotrophic food source (Fig. 8A-C):
lower N:C ratios (effectively representing lower DIN and/or higher
heterotrophy) favor nitrogen-limitation and more rapid recovery, while
higher N:C ratios (effectively representing higher DIN and/or lower
heterotrophy) favor carbon-limitation and prolonged recovery with a
larger overshoot in S:H biomass.

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Recovery from bleaching with varying N:C availability (A:
N:C=0.09; B: N:C=0.10; C: N:C=0.30). Higher N:C ratios in a
heterotrophic food source (effectively representing higher external DIN
and/or lower feeding rates) cause a larger overshoot in the S:H biomass
ratio, and prolong the duration of time until the system `recovers' by
re-establishing nitrogen-limitation. All simulations run with default
parameters (except varying nNX), with L=10 mol photons
m\textsuperscript{2} d\textsuperscript{-1}, DIN=1e-7 molN
L\textsuperscript{-1}, prey=1e-6 CmolX L\textsuperscript{-1}, and
initial S:H biomass=0.001.}
\end{figure}

These dynamics reveal that the re-establishment of nitrogen-limitation
is the most important diagnostic of recovery to a `healthy' state, as
this is when positive growth rates are resumed. A high S:H biomass ratio
alone does not necessarily indicate that a `healthy' state has been
reached, since the C-limited state may still persist (e.g., Fig. 8B,
8C). This could explain why corals that have recovered their symbiont
populations after bleaching may still exhibit energetic deficits and
physiological impacts for much longer periods of time (Levitan et al.
2014, Hughes and Grottoli (2013)). Therefore, host acquisition of carbon
from a source other than the symbiont may be extremely important for the
system to recover from bleaching. Indeed, host feeding has been shown to
promote a more rapid return to pre-bleaching levels of key physiological
parameters in recovering corals (Connolly, Lopez-Yglesias, and Anthony
2012). The model predicts that acquisition of carbon from other sources,
such as direct uptake of DOC (Levas et al. 2015), may also promote more
rapid recovery from a bleached state.

@@

\hypertarget{ref-Connolly:2012ep}{}
Connolly, Sean R, M A Lopez-Yglesias, and Kenneth R N Anthony. 2012.
``Food availability promotes rapid recovery from thermal stress in a
scleractinian coral.'' \emph{Coral Reefs} 31 (4). Springer-Verlag:
951--60.
doi:\href{https://doi.org/10.1007/s00338-012-0925-9}{10.1007/s00338-012-0925-9}.

@@

\hypertarget{ref-Hughes:2013hd}{}
Hughes, Adam D, and Andréa G Grottoli. 2013. ``Heterotrophic
Compensation: A Possible Mechanism for Resilience of Coral Reefs to
Global Warming or a Sign of Prolonged Stress?'' \emph{PLoS ONE} 8 (11).
Public Library of Science: e81172.
doi:\href{https://doi.org/10.1371/journal.pone.0081172}{10.1371/journal.pone.0081172}.

@@

\hypertarget{ref-Levas:2015ic}{}
Levas, Stephen, Andréa G Grottoli, Verena Schoepf, Matthew D
Aschaffenburg, Justin Baumann, James E Bauer, and Mark E Warner. 2015.
``Can heterotrophic uptake of dissolved organic carbon and zooplankton
mitigate carbon budget deficits in annually bleached corals?''
\emph{Coral Reefs} 35 (2). Springer Berlin Heidelberg: 495--506.
doi:\href{https://doi.org/10.1007/s00338-015-1390-z}{10.1007/s00338-015-1390-z}.

\hypertarget{ref-Levitan:2014et}{}
Levitan, D R, W Boudreau, J Jara, and Nancy Knowlton. 2014. ``Long-term
reduced spawning in Orbicella coral species due to temperature stress.''
\emph{Marine Ecology Progress Series} 515: 1--10.
doi:\href{https://doi.org/10.3354/meps11063}{10.3354/meps11063}.

@@

\hypertarget{ref-Wiedenmann:2013bn}{}
Wiedenmann, Jörg, C D'Angelo, Edward G Smith, Alan N Hunt, François-Eric
Legiret, Anthony D Postle, and Eric P Achterberg. 2013. ``Nutrient
enrichment can increase the susceptibility of reef corals to
bleaching.'' \emph{Nature Climate Change} 3. Nature Publishing Group:
160--64.
doi:\href{https://doi.org/10.1038/nclimate1661}{10.1038/nclimate1661}.
